Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-02-06 LaTeX/Turnover_plos.tex
trying to get figures into online system

Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}). 
Among species with genetic sex determination (GSD) of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes. 
For example, transitions of the master sex-determining locus between chromosomes or the evolution of new master sex-determining loci (where the heterogametic sex does not change, hereafter `cis-GSD transitions') have occurred in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), including snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}, amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

When only one neo-W haplotype has growth rate greater than one (see Fig \ref{fig:regionplots}), a neo-W allele can invade as long as equation \eqref{eq:lambdasGen} is satisfied, which may require that the recombination rate, $R$, is small enough.
Nevertheless, because we assume here that $r$ is small, these results indicate that a more loosely linked sex-determining region ($r<R$) can spread.
For example, tightly sex-linked loci that experience sexually-antagonistic selection can drive trans-GSD transitions in which the new sex-determining locus is less closely linked to the locus under selection (Fig \ref{fig:SexAntagTighter}).

@@

%The new sex-determining allele, $m$, will spread in an XY system if $\lambda_{m}^{(XY)}>1$. 
Equation \eqref{eq:lambda_neoY} demonstrates that, under weak selection, a neo-Y allele will invade an XY system ($\lambda_{Y'}^{(XY)}>1$) if and only if it is more closely linked to the selected locus than the ancestral sex-determining locus (i.e., if $R<r$; note that $V_{A}S_{A}^2$ is strictly positive as long as $\mathbf{A}$ is polymorphic). 
This echoes our results above where a neo-Y could never invade if $r\approx0$. 
It is also consistent with the results of \cite{vanDoorn:2007eu}, who considered diploid selection only and also found that cis-GSD transitions can only occur when the new sex-determining locus is more closely linked to a locus under sexually-antagonistic selection. 
%Because tighter linkage between the sex-determining region and a selected locus evolves during these transitions, stronger associations between selected alleles and diploid sexes can build up by selection (e.g., between male-beneficial alleles and the male-determining allele, Y).%, increasing diploid mean fitness (Fig \ref{fig:Combination_MeanFit}).

With weak selection and no haploid selection ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$), the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W'}^{(XY)}=\lambda_{Y'}^{(XY)}$), such that trans-GSD transitions can also occur only if the new sex-determining locus is more closely linked to a locus under selection ($R<r$), as found by \cite{vanDoorn:2010hu}. 
When there is haploid selection, invasion also typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Fig \ref{fig:Combination_Centimorgans}).
For example, if the $\mathbf{A}$ locus is unlinked to the ancestral sex-determining locus ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade.
In this case, there is no ancestral association between $\mathbf{A}$ alleles and sex chromosomes in males, $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$, see equation \eqref{eq:freq_diffs}. 
The second term in equation \eqref{eq:lambda_neoW} therefore disappears and invasion depends only on the sign of $(r-R)$.%, as in the case of the neo-Y.

@@

However, with haploid selection and $r<1/2$, the additional term in equation \eqref{eq:lambda_neoW} can be positive.
This can allow neo-W invasion ($\lambda_{W'}^{(XY)}>1$), even when the new sex-determining locus is less closely linked to the selected locus ($R>r$). 
%These transitions are unusual because, when $R>r$, associations that selection has built up between alleles more favourable in one sex and alleles that determine sex will be weakened. 
%Mean diploid fitness therefore decreases during heterogametic transitions that create looser sex-linkage (Fig \ref{fig:Combination_MeanFit}B,D). 
%Equation \eqref{eq:lambda_neoW} shows that, with weak selection, neo-W alleles can invade an XY system for a large number of selective regimes. 
%To clarify the parameter space under which $\lambda_{W'}^{(XY)}>1$ , we consider several special cases. 
%Firstly, if the $\mathbf{A}$ locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because there is no ancestral association between $A$ alleles and sex chromosomes in males, $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$, see equation \eqref{eq:freq_diffs}. 
%The second term in equation \eqref{eq:lambda_neoW} therefore disappears and invasion depends only on the sign of $(r-R)$, as in the case of the neo-Y. 
%Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Fig \ref{fig:Combination_Centimorgans}).
To clarify the parameter space under which invasion occurs despite looser sex-linkage ($\lambda_{W'}^{(XY)}>1$ despite $R>r$), we focus on the special case where $R=1/2$ and $r<1/2$ (e.g., the selected locus is on the ancestral sex chromosome and the novel sex-determining locus arises on an autosome). 
In Table \ref{tab:specialcases} we give the conditions where invasion occurs when we further assume that haploid selection only occurs in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
When there is no gametic competition and meiotic drive is in one sex only, an unlinked neo-W can invade as long as the same allele is favoured during diploid selection in males and females ($s^\female s^\male>0$, see Fig \ref{fig:Combination_Centimorgans}A and Fig \ref{fig:SexRatioBad}B). %, which is 50\% of the parameter space. 
When there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Fig \ref{fig:Combination_Centimorgans}B). %, which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion occurs for a relatively large fraction of the parameter space, even if the neo-W uncouples the sex-determining locus from a locus under selection.

@@

%Under Fisherian sex ratio selection, autosomal modifiers favour equal investment in male and female offspring, i.e., a 1:1 sex ratio~\cite{Fisher:1930wy,Charnov:1982wg,West:2009we}. 
%A novel environmental sex-determiner that causes half of its carriers to become female and half to become male ($k=1/2$) will be in males half of the time and in females half of the time (like an autosome).
%In addition, these novel sex-determination alleles equalize the sex ratio and therefore one might expect them to be favoured by Fisherian sex ratio selection when the resident sex ratio is biased.
%However, assuming weak selection, we find that the growth rate of a rare, dominant offspring-controlled neo-ESD allele that produces males or females with equal probability ($k=1/2$) is
Of particular interest are ESD mutations that cause half of their carriers to develop as females and half as males ($k=1/2$), creating equal sex ratios.
The spread of such ESD is determined by
\begin{equation}
\lambda_{ESD'}^{(XY)} =1+ \frac{1}{2}\frac{(\lambda_{Y'\rvert R=1/2}^{(XY)}-1) + (\lambda_{W'\rvert R=1/2}^{(XY)}-1)}{2} + O\left(\epsilon^3\right),
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s^\male - 3 s^\female\right) }{8 {\left( s_{f} + s_{m} \right)}^2} t_{m}^2.
\label{eq:lambda_ESD}
\end{equation}
\noindent
where $\lambda_{Y'\rvert R=1/2}^{(XY)}$ and $\lambda_{W'\rvert R=1/2}^{(XY)}$ represent $\lambda_{Y'}^{(XY)}$ and $\lambda_{W'}^{(XY)}$ when evaluated at $R=1/2$ (Equations \ref{eq:lambda_neoY} and \ref{eq:lambda_neoW}).
That is, recombination between the selected locus and the novel sex-determining locus, $R$, does not enter into the $k=1/2$ results.
This is because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-W and half the time as a neo-Y.
preventing associations from building up between alleles at locus $\mathbf{A}$ and sex. 
Equation \eqref{eq:lambda_ESD} shows that the ESD mutation gets half of the fitness of a feminizing mutation (neo-W) and half of the fitness of a masculinizing mutation (neo-Y), but only has an effect one half of the time (the other half of the time it produces the same sex as the ancestral system would have). 
As discussed above, $\lambda_{Y'\rvert R=1/2}^{(XY)}$ is necessarily less than one, but $\lambda_{W'\rvert R=1/2}^{(XY)}$ can be greater than one if there is haploid selection.
That is, when there is haploid selection, ESD can invade an ancestrally-XY system because it generates females that are either rare or have high fitness, in the same manner as a neo-W.

%An important result from equation \eqref{eq:lambda_ESD} is that ESD can invade if there is haploid selection. 
%When evaluated at $R=1/2$, $\Lambda_{Y',XY}$ is necessarily less than one, but $\Lambda_{W',XY}$ can be greater than one if there is haploid selection, as discussed above. 
%Previous studies where ESD is favoured have typically assumed that environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that ESD invades because it allows sex determination to depend on the environment \citep[reviewed in][]{Charnov:1982wg,Bull:1983vi,West:2009we}. 
%Here, ESD mutations can spread because they generate females that are either rare or have high fitness, in the same manner as a neo-W.

Significantly, equation \eqref{eq:lambda_ESD} is the same whether ESD is invading an ancestrally XY or ZW system (because $\lambda_{Y'}^{(XY)} = \lambda_{W'}^{(ZW)}$ and $\lambda_{W'}^{(XY)} = \lambda_{Y'}^{(ZW)}$).
Thus, Fisherian sex ratio selection alone does not easily explain GSD to ESD transitions.
For example, when the ancestral sex-determining system is XY, but not ZW, the sex ratio is biased by male haploid selection.
Nevertheless, ESD is equally likely to invade when it equalizes the zygotic sex ratio (through $\lambda_{W'}^{(XY)}$) as when it doesn't (through $\lambda_{Y'}^{(ZW)}$). 
%This is again because haploid selection can facilitate the turnover of sex chromosomes either through sex ratio biases or through associations between the new sex determining region with selected alleles. 
%With weak selection, these two forces are nearly equal in strength. 
In addition, we note that ESD may not invade, even if the sex ratio is initially biased (e.g., with drive in males only, $r<1/2$, $h^\female=h^\male$, and $s^\female s^\male<0$, then $\lambda_{W'}^{(XY)}<1$, see Table \ref{tab:specialcases}). 
We conclude that, as with neo-W and neo-Y loci, associations with selected loci mean that the evolution of ESD is not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present.

@@

We next calculate when equilibria $(A)$ and $(B)$ are locally stable for $r=0$. 
According to the `small parameter theory'~\cite{Karlin:1972ab,Karlin:1972dq}, these stability properties are unaffected by small amounts of recombination between the sex-determining locus and the $\mathbf{A}$ locus, although equilibrium frequencies may be slightly altered. 
For the $a$ allele to be stably fixed on the Y background we need $\bar{w}_{Ya}^{\male} > \bar{w}_{YA}^{\male}$ where $\bar{w}_{Ya}^{\male} = w_{a}^{\male} \big[\hat{p}_X^\female(1-\alpha^\male_\Delta) w_{A}^{\female} w_{Aa}^{\male} + (1-\hat{p}_X^\female)w_{a}^{\female} w_{aa}^{\male} \big]$ and $\bar{w}_{YA}^{\male} = w_{A}^{\male} \big[ \hat{p}_X^\female w_{A}^{\female} w_{AA}^{\male} + (1-\hat{p}_X^\female)(1+\alpha^\male_\Delta) w_{a}^{\female} w_{Aa}^{\male} \big]$. 
That is, Y-$a$ haplotypes must have higher fitness than Y-$A$ haplotypes.  
Substituting in $\hat{p}_X^\female$ from equation \eqref{eq:tightequil}, fixation of the $a$ allele on the Y background requires that $\gamma_{i}>0$ where $\gamma_{(A)}=w_{a}^{\male} \big[ (1-\alpha^\male_\Delta) w_{Aa}^{\male} \phi + w_{aa}^{\male} \psi \big]-w_{A}^{\male} \big[ w_{AA}^{\male} \phi + (1 + \alpha^{\male}_\Delta)w_{Aa}^{\male} \psi \big]$ for equilibrium $(A)$ and $\gamma_{(B)}=(1-\alpha^{\male}_\Delta)w_{a}^{\male}w_{Aa}^{\male}-w_{A}^{\male}w_{AA}^{\male}$ for equilibrium $(B)$.
Stability of a polymorphism on the X background (equilibrium $(A)$) further requires that $\phi >0$ and $\psi >0$. 
Fixation of the $a$ allele on the X background (equilibrium $(B)$) can be stable only if equilibrium $(A)$ is not, as it requires $\psi<0$. %and $2w_{A}^{\female}w_{AA}^{\female}>(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ or just $4w_{A}^{\female}w_{AA}^{\female}<(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ (which also prevents $\psi>0$).

@@

When $R=0$ the two roots are $\Lambda_{mA}^{(XY)}$ and $\Lambda_{ma}^{(XY)}$, and the leading eigenvalue is the larger of the two.
When $R>0$ then $f(\Lambda_{mA}^{(XY)})$ and $f(\Lambda_{ma}^{(XY)})$ are of opposite signs, and the leading eigenvalue must fall between these two quantities (details in \nameref{file:Mathematica}). %(the Perron-Frobenius theorem guarantees that the largest eigenvalue is positive, unique, and real).  
Thus, $\lambda_m^{(XY)}>1$ if both $\Lambda_{mA}^{(XY)}>1$ and $\Lambda_{ma}^{(XY)}>1$; similarly, $\lambda_m^{(XY)}<1$ if both $\Lambda_{mA}^{(XY)}<1$ and $\Lambda_{ma}^{(XY)}<1$.
If only one haplotypic growth rate is greater than one ($\Lambda_{ma}^{(XY)}<1<\Lambda_{mA}^{(XY)}$ or $\Lambda_{mA}^{(XY)}<1<\Lambda_{ma}^{(XY)}$), then $\lambda_m^{(XY)}$ is greater than one when condition \eqref{eq:lambdasGen} is met. 
Thus, the invasion of a new sex-determining allele is determined by the haplotypic growth rates ($\Lambda_{mj}^{(XY)}$ terms), which do not account for loss due to recombination, and the rates at which recombination between the $\mathbf{M}$ and $\mathbf{A}$ loci breaks apart these haplotypes ($\chi_{mj}^{(XY)}$).
For tight linkage between the ancestral sex-determining locus ($\mathbf{X}$) and the selected locus ($\mathbf{A}$) we can calculate these terms explicitly (see below).
For weak selection we approximate the leading eigenvalue with a Taylor series. 
The leading eigenvalue for any $k$ is given up to order $\epsilon^2$ by equation \eqref{eq:lambda_ESD_k}.

>>>>

Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}). 
Among species with genetic sex determination (GSD) of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes. 
For example, transitions of the master sex-determining locus between chromosomes or the evolution of new master sex-determining loci where the heterogametic sex does not change (hereafter `cis-GSD transitions') have occurred in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), including snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}, amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

When only one neo-W haplotype has growth rate greater than one (see Fig \ref{fig:regionplots}), a neo-W allele can invade as long as Eq \eqref{eq:lambdasGen} is satisfied, which may require that the recombination rate, $R$, is small enough.
Nevertheless, because we assume here that $r$ is small, these results indicate that a more loosely linked sex-determining region ($r<R$) can spread.
For example, tightly sex-linked loci that experience sexually-antagonistic selection can drive trans-GSD transitions in which the new sex-determining locus is less closely linked to the locus under selection (Fig \ref{fig:SexAntagTighter}).

@@

%The new sex-determining allele, $m$, will spread in an XY system if $\lambda_{m}^{(XY)}>1$. 
Eq \eqref{eq:lambda_neoY} demonstrates that, under weak selection, a neo-Y allele will invade an XY system ($\lambda_{Y'}^{(XY)}>1$) if and only if it is more closely linked to the selected locus than the ancestral sex-determining locus (i.e., if $R<r$; note that $V_{A}S_{A}^2$ is strictly positive as long as $\mathbf{A}$ is polymorphic). 
This echoes our results above where a neo-Y could never invade if $r\approx0$. 
It is also consistent with the results of \cite{vanDoorn:2007eu}, who considered diploid selection only and also found that cis-GSD transitions can only occur when the new sex-determining locus is more closely linked to a locus under sexually-antagonistic selection. 
%Because tighter linkage between the sex-determining region and a selected locus evolves during these transitions, stronger associations between selected alleles and diploid sexes can build up by selection (e.g., between male-beneficial alleles and the male-determining allele, Y).%, increasing diploid mean fitness (Fig \ref{fig:Combination_MeanFit}).

With weak selection and no haploid selection ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$), the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W'}^{(XY)}=\lambda_{Y'}^{(XY)}$), such that trans-GSD transitions can also occur only if the new sex-determining locus is more closely linked to a locus under selection ($R<r$), as found by \cite{vanDoorn:2010hu}. 
When there is haploid selection, invasion also typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Fig \ref{fig:Combination_Centimorgans}).
For example, if the $\mathbf{A}$ locus is unlinked to the ancestral sex-determining locus ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade.
In this case, there is no ancestral association between $\mathbf{A}$ alleles and sex chromosomes in males, $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$, see Eq \eqref{eq:freq_diffs}. 
The second term in Eq \eqref{eq:lambda_neoW} therefore disappears and invasion depends only on the sign of $(r-R)$.%, as in the case of the neo-Y.

@@

However, with haploid selection and $r<1/2$, the additional term in Eq \eqref{eq:lambda_neoW} can be positive.
This can allow neo-W invasion ($\lambda_{W'}^{(XY)}>1$), even when the new sex-determining locus is less closely linked to the selected locus ($R>r$). 
%These transitions are unusual because, when $R>r$, associations that selection has built up between alleles more favourable in one sex and alleles that determine sex will be weakened. 
%Mean diploid fitness therefore decreases during heterogametic transitions that create looser sex-linkage (Fig \ref{fig:Combination_MeanFit}B,D). 
%Eq \eqref{eq:lambda_neoW} shows that, with weak selection, neo-W alleles can invade an XY system for a large number of selective regimes. 
%To clarify the parameter space under which $\lambda_{W'}^{(XY)}>1$ , we consider several special cases. 
%Firstly, if the $\mathbf{A}$ locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because there is no ancestral association between $A$ alleles and sex chromosomes in males, $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$, see Eq \eqref{eq:freq_diffs}. 
%The second term in Eq \eqref{eq:lambda_neoW} therefore disappears and invasion depends only on the sign of $(r-R)$, as in the case of the neo-Y. 
%Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Fig \ref{fig:Combination_Centimorgans}).
To clarify the parameter space under which invasion occurs despite looser sex-linkage ($\lambda_{W'}^{(XY)}>1$ despite $R>r$), we focus on the special case where $R=1/2$ and $r<1/2$ (e.g., the selected locus is on the ancestral sex chromosome and the novel sex-determining locus arises on an autosome). 
In Table \ref{tab:specialcases} we give the conditions where invasion occurs when we further assume that haploid selection only occurs in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
When there is no gametic competition and meiotic drive is in one sex only, an unlinked neo-W can invade as long as the same allele is favoured during diploid selection in males and females ($s^\female s^\male>0$, see Fig \ref{fig:Combination_Centimorgans}A and Fig \ref{fig:SexRatioBad}B). %, which is 50\% of the parameter space. 
When there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Fig \ref{fig:Combination_Centimorgans}B). %, which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion occurs for a relatively large fraction of the parameter space, even if the neo-W uncouples the sex-determining locus from a locus under selection.

@@

%Under Fisherian sex ratio selection, autosomal modifiers favour equal investment in male and female offspring, i.e., a 1:1 sex ratio~\cite{Fisher:1930wy,Charnov:1982wg,West:2009we}. 
%A novel environmental sex-determiner that causes half of its carriers to become female and half to become male ($k=1/2$) will be in males half of the time and in females half of the time (like an autosome).
%In addition, these novel sex-determination alleles equalize the sex ratio and therefore one might expect them to be favoured by Fisherian sex ratio selection when the resident sex ratio is biased.
%However, assuming weak selection, we find that the growth rate of a rare, dominant offspring-controlled neo-ESD allele that produces males or females with equal probability ($k=1/2$) is
Of particular interest are ESD mutations that cause half of their carriers to develop as females and half as males ($k=1/2$), creating equal sex ratios.
The spread of such ESD is determined by
\begin{equation}
\lambda_{ESD'}^{(XY)} =1+ \frac{1}{2}\frac{(\lambda_{Y'\rvert R=1/2}^{(XY)}-1) + (\lambda_{W'\rvert R=1/2}^{(XY)}-1)}{2} + O\left(\epsilon^3\right),
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s^\male - 3 s^\female\right) }{8 {\left( s_{f} + s_{m} \right)}^2} t_{m}^2.
\label{eq:lambda_ESD}
\end{equation}
\noindent
where $\lambda_{Y'\rvert R=1/2}^{(XY)}$ and $\lambda_{W'\rvert R=1/2}^{(XY)}$ represent $\lambda_{Y'}^{(XY)}$ and $\lambda_{W'}^{(XY)}$ when evaluated at $R=1/2$ (Equations \ref{eq:lambda_neoY} and \ref{eq:lambda_neoW}).
That is, recombination between the selected locus and the novel sex-determining locus, $R$, does not enter into the $k=1/2$ results.
This is because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-W and half the time as a neo-Y.
preventing associations from building up between alleles at locus $\mathbf{A}$ and sex. 
Eq \eqref{eq:lambda_ESD} shows that the ESD mutation gets half of the fitness of a feminizing mutation (neo-W) and half of the fitness of a masculinizing mutation (neo-Y), but only has an effect one half of the time (the other half of the time it produces the same sex as the ancestral system would have). 
As discussed above, $\lambda_{Y'\rvert R=1/2}^{(XY)}$ is necessarily less than one, but $\lambda_{W'\rvert R=1/2}^{(XY)}$ can be greater than one if there is haploid selection.
That is, when there is haploid selection, ESD can invade an ancestrally-XY system because it generates females that are either rare or have high fitness, in the same manner as a neo-W.

%An important result from Eq \eqref{eq:lambda_ESD} is that ESD can invade if there is haploid selection. 
%When evaluated at $R=1/2$, $\Lambda_{Y',XY}$ is necessarily less than one, but $\Lambda_{W',XY}$ can be greater than one if there is haploid selection, as discussed above. 
%Previous studies where ESD is favoured have typically assumed that environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that ESD invades because it allows sex determination to depend on the environment \citep[reviewed in][]{Charnov:1982wg,Bull:1983vi,West:2009we}. 
%Here, ESD mutations can spread because they generate females that are either rare or have high fitness, in the same manner as a neo-W.

Significantly, Eq \eqref{eq:lambda_ESD} is the same whether ESD is invading an ancestrally XY or ZW system (because $\lambda_{Y'}^{(XY)} = \lambda_{W'}^{(ZW)}$ and $\lambda_{W'}^{(XY)} = \lambda_{Y'}^{(ZW)}$).
Thus, Fisherian sex ratio selection alone does not easily explain GSD to ESD transitions.
For example, when the ancestral sex-determining system is XY, but not ZW, the sex ratio is biased by male haploid selection.
Nevertheless, ESD is equally likely to invade when it equalizes the zygotic sex ratio (through $\lambda_{W'}^{(XY)}$) as when it doesn't (through $\lambda_{Y'}^{(ZW)}$). 
%This is again because haploid selection can facilitate the turnover of sex chromosomes either through sex ratio biases or through associations between the new sex determining region with selected alleles. 
%With weak selection, these two forces are nearly equal in strength. 
In addition, we note that ESD may not invade, even if the sex ratio is initially biased (e.g., with drive in males only, $r<1/2$, $h^\female=h^\male$, and $s^\female s^\male<0$, then $\lambda_{W'}^{(XY)}<1$, see Table \ref{tab:specialcases}). 
We conclude that, as with neo-W and neo-Y loci, associations with selected loci mean that the evolution of ESD is not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present.

@@

We next calculate when equilibria $(A)$ and $(B)$ are locally stable for $r=0$. 
According to the `small parameter theory'~\cite{Karlin:1972ab,Karlin:1972dq}, these stability properties are unaffected by small amounts of recombination between the sex-determining locus and the $\mathbf{A}$ locus, although equilibrium frequencies may be slightly altered. 
For the $a$ allele to be stably fixed on the Y background we need $\bar{w}_{Ya}^{\male} > \bar{w}_{YA}^{\male}$ where $\bar{w}_{Ya}^{\male} = w_{a}^{\male} \big[\hat{p}_X^\female(1-\alpha^\male_\Delta) w_{A}^{\female} w_{Aa}^{\male} + (1-\hat{p}_X^\female)w_{a}^{\female} w_{aa}^{\male} \big]$ and $\bar{w}_{YA}^{\male} = w_{A}^{\male} \big[ \hat{p}_X^\female w_{A}^{\female} w_{AA}^{\male} + (1-\hat{p}_X^\female)(1+\alpha^\male_\Delta) w_{a}^{\female} w_{Aa}^{\male} \big]$. 
That is, Y-$a$ haplotypes must have higher fitness than Y-$A$ haplotypes.  
Substituting in $\hat{p}_X^\female$ from Eq \eqref{eq:tightequil}, fixation of the $a$ allele on the Y background requires that $\gamma_{i}>0$ where $\gamma_{(A)}=w_{a}^{\male} \big[ (1-\alpha^\male_\Delta) w_{Aa}^{\male} \phi + w_{aa}^{\male} \psi \big]-w_{A}^{\male} \big[ w_{AA}^{\male} \phi + (1 + \alpha^{\male}_\Delta)w_{Aa}^{\male} \psi \big]$ for equilibrium $(A)$ and $\gamma_{(B)}=(1-\alpha^{\male}_\Delta)w_{a}^{\male}w_{Aa}^{\male}-w_{A}^{\male}w_{AA}^{\male}$ for equilibrium $(B)$.
Stability of a polymorphism on the X background (equilibrium $(A)$) further requires that $\phi >0$ and $\psi >0$. 
Fixation of the $a$ allele on the X background (equilibrium $(B)$) can be stable only if equilibrium $(A)$ is not, as it requires $\psi<0$. %and $2w_{A}^{\female}w_{AA}^{\female}>(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ or just $4w_{A}^{\female}w_{AA}^{\female}<(1-\alpha_\Delta^{\female})w_{a}^{\female}w_{Aa}^{\female}$ (which also prevents $\psi>0$).

@@

When $R=0$ the two roots are $\Lambda_{mA}^{(XY)}$ and $\Lambda_{ma}^{(XY)}$, and the leading eigenvalue is the larger of the two.
When $R>0$ then $f(\Lambda_{mA}^{(XY)})$ and $f(\Lambda_{ma}^{(XY)})$ are of opposite signs, and the leading eigenvalue must fall between these two quantities (details in \nameref{file:Mathematica}). %(the Perron-Frobenius theorem guarantees that the largest eigenvalue is positive, unique, and real).  
Thus, $\lambda_m^{(XY)}>1$ if both $\Lambda_{mA}^{(XY)}>1$ and $\Lambda_{ma}^{(XY)}>1$; similarly, $\lambda_m^{(XY)}<1$ if both $\Lambda_{mA}^{(XY)}<1$ and $\Lambda_{ma}^{(XY)}<1$.
If only one haplotypic growth rate is greater than one ($\Lambda_{ma}^{(XY)}<1<\Lambda_{mA}^{(XY)}$ or $\Lambda_{mA}^{(XY)}<1<\Lambda_{ma}^{(XY)}$), then $\lambda_m^{(XY)}$ is greater than one when condition \eqref{eq:lambdasGen} is met. 
Thus, the invasion of a new sex-determining allele is determined by the haplotypic growth rates ($\Lambda_{mj}^{(XY)}$ terms), which do not account for loss due to recombination, and the rates at which recombination between the $\mathbf{M}$ and $\mathbf{A}$ loci breaks apart these haplotypes ($\chi_{mj}^{(XY)}$).
For tight linkage between the ancestral sex-determining locus ($\mathbf{X}$) and the selected locus ($\mathbf{A}$) we can calculate these terms explicitly (see below).
For weak selection we approximate the leading eigenvalue with a Taylor series. 
The leading eigenvalue for any $k$ is given up to order $\epsilon^2$ by Eq \eqref{eq:lambda_ESD_k}.
